Titeln på avhandlingens språk

### 2022-11-22 body/part1/overview.tex
edits

\section{Methodology}

\todo[inline]{
    Study by example, human subject research, and modeling.

@@

    Add a hardened compilation of key results, conclusions, figures, and tables.
    Synthesize and summarize the research questions and conclusions to find the red thread.

    EdgeDroid (first demo paper at SEC, then HotMobile paper).

    Tie in CLEAVE paper here as an extension of the approach towards other latency-sensitive applications.

    Potentially include Ainur paper as a further extension of the methodology?

    PLOS ONE paper as a stepping stone towards the final goal of a more realistic model?

    Finally, EdgeDroid2 paper:
    Presents the final model of human behavior
    Applies it to a real optimization problem, employing theoretical results proved by Vishnu
}

We summarize the seven key contributions of this thesis below:

\begin{enumerate}
    \item\label{item:contrib:methodology} A methodological approach to studying system responsiveness versus resource consumption trade-offs in \ac{WCA}.
    This approach is based on the emulation of the client-side workload, while maintaining the \emph{real} server-side process as well as network stack.
    \item In conjunction with the aforementioned methodology, a tool (EdgeDroid) for the emulation of human behavior in \ac{WCA}.
    This tool implements the first ever parametrizable model of human behavior for these applications using empirical data collected from a series of human-subject experiments.
    \item A characterization of the effects of individual personality differences in human responses to reduced system responsiveness in interactive systems such as \acl{WCA}.
    In particular, this thesis presents novel evidence for the modulating properties of \emph{neuroticism} on the reaction to increased delay in these systems.
    \item An preliminary, yet novel exploration of the implications of realistic modeling of human timings in \ac{WCA} applications for their optimization potential.
    \item\label{item:contrib:methodology:ncs} An extension of the methodology described in \cref{item:contrib:methodology} towards \acp{NCS}.
    \item A novel tool, CLEAVE, for the emulation of \acp{NCS} on edge infrastructure.
    This tool has been designed from the ground-up to be edge-native, implemented using widely-adopted technologies in the disciplines of cloud and edge computing.
    \item Finally, a software framework for the management and automation of the testbeds required for the methodology described in \cref{item:contrib:methodology,item:contrib:methodology:ncs}.
\end{enumerate}

@@

\todo[inline]{
    \cref{paper:EdgeDroid1.0} introduces our methodology, and validates its use in studying WCA.
    Focus less on EdgeDroid 1.0 and more on the methodology, although discuss that EdgeDroid 1.0 is an initial approach to a tool we complete with EdgeDroid 2.0.
}

@@

\todo[inline]{PLOS paper describes the acquisition of the necessary data and insights for the models.}

We still have a very limited understanding of how humans react to delays in these systems --- specifically, how changes in system responsiveness impact users.
\emph{System responsiveness} here denotes a qualitative scale ranging from \emph{high} (that is, not subject to delay or subject to negligible delay with respect to human perception) to \emph{low} (i.e.\ subject to considerable delay).
% Conversely, QoE is understood to range from \emph{good} (i.e.\ users enjoy interacting with the application) to bad (users dislike interacting with the application).
Characterizing the relationships between system responsiveness and user behavior and quality of experience is of paramount importance for the design and evaluation of these applications.
It is generally acknowledged, for instance, that a system going from a state of high responsiveness to one of low responsiveness can cause a drop in quality of experience~\cite{dabrowsky:2011:40years}. 
Furthermore, this could cause users to modify the temporal paramaters of their behavior when interacting with the system, generating a sort of \emph{feedback loop} between user and system.
A clear understanding of these relationships would allow, for instance, for the development of strategies for load balancing and optimization for large-scale deployment of WCA.\@

This paper builds upon preliminary work in the field of time perception and delay characterization of WCA.\@
We expand upon the findings of Ha et al.~\cite{Ha:TowardsWearableCogAssist}, who identified the need for low-latency offloading in WCA, and of Chen et al.~\cite{Chen:AnEmpiricalStudyOfLatency}, who outlined the bounds for ``noticeable'' and ``unbearable'' latencies in these systems.
While these bounds present a general understanding of when it is likely that a user will stop using an application, they do not provide any insights as to what happens \emph{before} that --- i.e.\ how human behavior changes with system responsiveness.
We aim to tackle this gap in knowledge through the characterization of human responses to delays in the application pipeline, using latencies in the range defined by the previously established bounds.
This is an important step toward a more systematic understanding of human behavior in this domain.

We present in this paper the design and elaboration of an experimental WCA test-bed.
This test-bed was subsequently employed in a study in which undergraduate students of diverse fields of study, aged between 18 and 25 years old, were asked to interact with and follow the instructions given to them by a cognitive assistant.
Unbeknownst to the participants, we altered the responsiveness of the system in real-time and recorded the resulting behavioral and physiological reactions. 
The participants wore an array of biometric sensors measuring physiological responses that have proven useful in assessing cognitive workload during human-computer interaction~\cite{haapalainen2010psycho,kumar2016measurement} such as heart rate and EEG.\@

Through this experimental set-up, we intended to answer four core research questions relating to human responses to decreased application responsiveness.

@@

  The results we present, however, seem to refute this hypothesis. 
  We were not able to detect any significant effects on the physiological signals obtained from the biometric sensors as sytem responsiveness was altered.

@@

@@

\subsubsection{Implementing the model}

In that context, the contributions of this paper are as follows:
\begin{enumerate}
    \item\label{item:contrib:model} We introduce the first, to our knowledge, data-driven model for human timings in \ac{WCA} applications.
    Using the data collected for \textcite{olguinmunoz:impact2021} as a base, we build a stochastic model which takes as input past measurements of system responsiveness and produces realistic step execution times.
    We also introduce a novel way to generate dynamic traces of frames for \ac{WCA} applications which can be combined with the timing model for a full end-to-end emulation of a human.
    We name this new model \emph{EdgeDroid 2.0}; a direct, more realistic evolution of our initial EdgeDroid approach~\cite{olguin2018scaling,olguin2019edgedroid}.
    \item\label{item:contrib:footprint} Using the model described above, we study the implications of realistic human behavior on the application lifetime footprint of \ac{WCA}.
    In concordance with previous work~\cite{olguinmunoz:impact2021}, we find dependencies between system responsiveness and human step execution times that lead to substantially different application lifetimes when compared to a first-order baseline which does not take into account human behavior.
    \item\label{item:contrib:optimization} Finally, we study the potential for optimization in \ac{WCA} when considering human behavior using our model.
    We develop a generic model for the stochastic optimization of resource consumption versus responsiveness trade-offs in these applications, which we apply to two potential avenues for \ac{WCA} optimization; number of processed samples and energy consumption per step.
    Compared to the state-of-the-art, our optimization model results in up to \textasciitilde\SI{60}{\percent} fewer samples processed or an average of \SI{20}{\percent} less energy consumed per step, while maintaining comparable levels of responsiveness.
\end{enumerate}

@@

\subsubsection{Implications of the EdgeDroid model for \ac{WCA} optimization}

@@

\section{A methodology for the study of latency trade-offs in \acsp{NCS}}

@@

\section{The Ainur software framework for Edge Computing testbeds}

@@

>>>>

\todo[inline]{
    Methodology

    Study by example, human subject research, and modeling.

@@

    Add a hardened compilation of key results, conclusions, figures, and tables.
    Synthesize and summarize the research questions and conclusions to find the red thread.
}

\begin{contributions}
    \begin{enumerate}[
            before = \leavevmode%\vspace{-.5\baselineskip}
        ]
        \item\label{item:contrib:methodology} A methodological approach to studying system responsiveness versus resource consumption trade-offs in \ac{WCA}.
        This approach is based on the emulation of the client-side workload, while maintaining the \emph{real} server-side process as well as network stack.
        \item In conjunction with the aforementioned methodology, a tool (EdgeDroid) for the emulation of human behavior in \ac{WCA}.
              This tool implements the first ever parametrizable model of human behavior for these applications using empirical data collected from a series of human-subject experiments.
        \item A characterization of the effects of individual personality differences in human responses to reduced system responsiveness in interactive systems such as \acl{WCA}.
              In particular, this thesis presents novel evidence for the modulating properties of \emph{neuroticism} on the reaction to increased delay in these systems.
        \item An preliminary, yet novel exploration of the implications of realistic modeling of human timings in \ac{WCA} applications for their optimization potential.
              \item\label{item:contrib:methodology:ncs} An extension of the methodology described in \cref{item:contrib:methodology} towards \acp{NCS}.
        \item A novel tool, CLEAVE, for the emulation of \acp{NCS} on edge infrastructure.
              This tool has been designed from the ground-up to be edge-native, implemented using widely-adopted technologies in the disciplines of cloud and edge computing.
        \item Finally, a software framework for the management and automation of the testbeds required for the methodology described in \cref{item:contrib:methodology,item:contrib:methodology:ncs}.
    \end{enumerate}
\end{contributions}

@@

\todo[inline]{
    \cref{paper:Olguin:DemoScaling2018,paper:Olguin:EdgeDroid2019} introduce our methodology, and validate its use in studying WCA.
    Focus less on EdgeDroid 1.0 and more on the methodology, although discuss that EdgeDroid 1.0 is an initial approach to a tool we complete with EdgeDroid 2.0.
}

@@

\todo[inline]{\cref{paper:Olguin:ImpactWCA2021} describes the acquisition of the necessary data and insights for the models.}

\todo[inline]{From introduction}

We intended to answer four core research questions relating to human responses to decreased application responsiveness.

@@

          The results we present, however, seem to refute this hypothesis.
          We were not able to detect any significant effects on the physiological signals obtained from the biometric sensors as system responsiveness was altered.

@@

\todo[inline]{From discussion}

@@

\todo[inline]{Some diagrams, plots, and results}

\begin{figure}[h]
    \centering
    \begin{subfigure}[t]{.49\textwidth}
        \centering
        \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4a.eps}
        \caption{Structure of a step in a generic cognitive assistance application.
            The assistant provides an instruction to the user and continuously samples the step state; inputs captured while the step is unfinished are silently ``discarded'' (i.e.\ they do not cause the generation of a new instruction).
            % by the assistant, as they do not contain relevant information.
            Once the user finishes performing the step, the next sample \emph{will} cause the generation of a new instruction.
            % which will subsequently be provided to the user.}%
        }
        \label{fig:cogassist:step}
    \end{subfigure}%
    % \medskip%
    \hfill%
    \begin{subfigure}[t]{.49\textwidth}
        \centering
        \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4b.eps}
        \caption{In the experimental task, an additional variable segment of time is introduced immediately following the processing of the input frame in order to extend the perceived processing time of the input to a specific target delay.}%
        \label{fig:cogassist:step:delay}
    \end{subfigure}
    \medskip%
    \begin{subfigure}[t]{.49\textwidth}
        \centering
        \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4c.eps}
        \caption{Structure of a block in the experimental task.}%
        \label{fig:cogassist:block}
    \end{subfigure}%
    % \medskip%
    \hfill%
    \begin{subfigure}[t]{.49\textwidth}
        \centering
        \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4d.eps}
        \caption{Visualization of the execution time of a step.}%
        \label{fig:exectime:diagram}
    \end{subfigure}%
    % \medskip%
    \caption{Components of the cognitive assistance task.}
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig6.eps}
    \caption{Per-step execution time by block length vs.\ delay. Error bars indicate the Standard Error of the Mean (S.E.M.)}\label{fig:exectime}%    
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig7.eps}
    \caption{Mean per-step execution time vs.\ delay, by step slice.
        Error bars indicate S.E.M.}
    \label{fig:exectime:delay:slice}
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[width=.8\textwidth]{publications/2021ImpactDelayedResponse/Fig8.eps}
    \caption{Per-step execution time across the first four steps after a block transition from block \( B_{k-1} \) to \( B_k \). Error bars indicate S.E.M.}\label{fig:exectime:transition}%
\end{figure}

\begin{table}[h]
    \centering
    \caption{Principal Component Analysis}\label{tab:pca}
    \begin{subtable}[h]{\textwidth}
        \centering
        \caption{Main components identified.}
        \setlength{\tabcolsep}{0pt} % let TeX compute the intercolumn space
        \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}lrrr@{}}
            \toprule
            \textbf{Factor} & \textbf{Comp. 1} & \textbf{Comp. 2} & \textbf{Comp. 3} \\
            \midrule
            BFI Conscientiousness                  & \textcolor{lightgray}{\( -0.022 \)} &                         \( 0.668 \) & \textcolor{lightgray}{\( -0.481 \)} \\
            BFI Neuroticism                        &                         \( 0.600 \) &                        \( -0.678 \) & \textcolor{lightgray}{\( -0.118 \)} \\
            ITQ Focus                              &                         \( 0.678 \) &  \textcolor{lightgray}{\( 0.203 \)} &                         \( 0.504 \) \\
            ITQ Involvement                        &                         \( 0.573 \) &                         \( 0.540 \) &  \textcolor{lightgray}{\( 0.417 \)} \\
            Exec. Time (delay 3.0 s, length 12)    &                         \( 0.758 \) & \textcolor{lightgray}{\( -0.178 \)} & \textcolor{lightgray}{\( -0.348 \)} \\
            Log EEG power \( \alpha + \beta \) Slice 1 & \textcolor{lightgray}{\( -0.436 \)} & \textcolor{lightgray}{\( -0.251 \)} &                         \( 0.589 \) \\
            \bottomrule
        \end{tabular*}
    \end{subtable}
    \newline
    \medskip
    \newline
    \begin{subtable}[h]{\textwidth}
        \centering
        \caption{Percentages of variance explained by the components.}
        \begin{tabular*}{\columnwidth}{@{\extracolsep{\fill}\quad}lrrrr@{}}
            \toprule
            {} & \textbf{Comp. 1} & \textbf{Comp. 2} & \textbf{Comp. 3} & \textbf{Total} \\
            \midrule
            Explained Variance & \SI{31.88}{\percent} & \SI{22.22}{\percent} & \SI{19.03}{\percent} & \SI{73.13}{\percent} \\
            \bottomrule
        \end{tabular*}
    \end{subtable}
\end{table}

\FloatBarrier%
\subsubsection{Implementing the model}

\todo[inline]{Present the implementation of the model from \cref{paper:EdgeDroid2}}

We introduce the first, to our knowledge, data-driven model for human timings in \ac{WCA} applications.
Using the data collected for \textcite{olguinmunoz:impact2021} as a base, we build a stochastic model which takes as input past measurements of system responsiveness and produces realistic step execution times.
We also introduce a novel way to generate dynamic traces of frames for \ac{WCA} applications which can be combined with the timing model for a full end-to-end emulation of a human.
We name this new model \emph{EdgeDroid 2.0}; a direct, more realistic evolution of our initial EdgeDroid approach~\cite{olguin2018scaling,olguin2019edgedroid}.

@@

\todo[inline]{TODO: Some relevant results. Need to add paper to thesis.}

\subsection{Implications of the methodology for \ac{WCA} optimization}

\todo[inline]{TODO, same as above}
\todo[inline]{Include discussion from EdgeDroid 1.0?}

\begin{enumerate}
    \item\label{item:contrib:footprint} Using the model described above, we study the implications of realistic human behavior on the application lifetime footprint of \ac{WCA}.
    In concordance with previous work~\cite{olguinmunoz:impact2021}, we find dependencies between system responsiveness and human step execution times that lead to substantially different application lifetimes when compared to a first-order baseline which does not take into account human behavior.
    \item\label{item:contrib:optimization} Finally, we study the potential for optimization in \ac{WCA} when considering human behavior using our model.
    We develop a generic model for the stochastic optimization of resource consumption versus responsiveness trade-offs in these applications, which we apply to two potential avenues for \ac{WCA} optimization; number of processed samples and energy consumption per step.
    Compared to the state-of-the-art, our optimization model results in up to \textasciitilde\SI{60}{\percent} fewer samples processed or an average of \SI{20}{\percent} less energy consumed per step, while maintaining comparable levels of responsiveness.
\end{enumerate}

@@

\FloatBarrier%
\section{A methodology for the study of latency trade-offs in \acsp{NCS}}

\todo[inline]{\Cref{paper:CLEAVE2022} implements both the methodology and the tool.}

@@

\todo[inline]{Some plots and results}

\begin{figure}
    \centering
    \includegraphics[width=.6\columnwidth]{publications/2022CLEAVE/images/inverted_pendulum.png}
    \caption{
        The 2D inverted pendulum system.
    }\label{fig:invpend}
\end{figure}

\begin{figure}[t]
    \centering
    \begin{subfigure}[h]{.22\textwidth}
        \centering
        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_topple_frac}
        \caption{Toppled plants.}\label{fig:video:toppled}
    \end{subfigure}%
    \hfill%
    \begin{subfigure}[h]{.22\textwidth}
        \centering
        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_angle_rms}
        \caption{Angle \ac{RMS}.}\label{fig:video:rms}
    \end{subfigure}\\
    \begin{subfigure}[h]{.22\textwidth}
        \centering
        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_drop_frac}
        \caption{Packet losses.}\label{fig:video:drop}
    \end{subfigure}%
    \hfill%
    \begin{subfigure}[h]{.22\textwidth}
        \centering
        \includegraphics[width=\textwidth]{publications/2022CLEAVE/plots/fixed_video_rtt}
        \caption{\acsp{RTT}.}\label{fig:video:rtt}
    \end{subfigure}%
    \caption{
        Multi-loop, resource-constrained setup results.
        Error bars indicate \SI{95}{\percent} \acp{CI} in all plots.
    }\label{fig:video:results}
\end{figure}

\FloatBarrier%
\section{A software framework for Edge-computing testbeds}

\todo[inline]{\cref{paper:Ainur2022} describes a software framework for testbeds used for the methodology described in this thesis}

@@

\todo[inline]{Some plots}

\begin{figure}[t]
    \centering
    \includegraphics[width=0.9\linewidth]{publications/2022Ainur/figures/overview.jpg}
    \caption{Layered structure of an Ainur experiment}\label{fig:overview}
\end{figure}

\begin{figure*}
    \centering
    \includegraphics[width=0.8\textwidth]{publications/2022Ainur/figures/flow2.png}
    \caption{Lifecycle of an experimental run in Ainur. Blocks in orange are optional.}\label{fig:flow}
\end{figure*}

\begin{figure*}
    \centering
    \includegraphics[width=.8\textwidth]{publications/2022Ainur/figures/network.png}
    \caption{Network structure assumed by Ainur.}\label{fig:network}
\end{figure*}
